Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2018-10-26 mprop/mprop.tex
Small changes

% TODO: table of notation?
\newcommand{\Kuu}{\mathbf{K}_{\mathbf{u},\mathbf{u}}}
\newcommand{\Krr}{\mathbf{K}_{\mathbf{r},\mathbf{r}}}
\newcommand{\Kru}{\mathbf{K}_{\mathbf{r},\mathbf{u}}}
\newcommand{\DKL}{D_{\mathrm{KL}}}
\newcommand{\pfull}{p(\mathcal{D}, \Theta, \mathbf{X_u}, \mathbf{u}, \mathbf{r})}
\newcommand{\pfullS}{p(\mathcal{D}, \Theta_s, \mathbf{X_u}, \mathbf{u}_s, \mathbf{r}_s)}
\newcommand{\posterior}{p(\Theta, \mathbf{u}, \mathbf{r} | \mathcal{D}, \mathbf{X_u})}
\newcommand{\approximation}{q_{\bm\nu}(\Theta, \mathbf{u}, \mathbf{r})}
\newcommand{\approximationS}{q_{\bm\nu}(\Theta_s, \mathbf{u}_s, \mathbf{r}_s)}
\newcommand{\Eq}{\mathbb{E}_{(\Theta, \mathbf{u}, \mathbf{r}) \sim \approximation}}

@@

The likelihood of the data can be written down as
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}
\begin{equation} \label{pDr}
  p(\mathcal{D} | r) = \prod_{i=1}^N \prod_{t=1}^T p(a_{i,t} | s_{i,t}) = \exp\left( \sum_{i=1}^N \sum_{t=1}^T Q(s_{i,t}, a_{i,t}; r) - V(s_{i,t}; r) \right),
\end{equation}
where
\[ Q(s_{i,t}, a_{i,t}; r) = r(s_{i,t}) + \gamma\sum_{s' \in \mathcal{S}}
  \mathcal{T}(s_{i,t}, a_{i,t}, s')V(s'; r), \]
and $V(s; r)$ can be obtained by repeatedly applying the equation
\cite{supplementary_material}
\[ V(s; r) = \log \sum_{a \in \mathcal{A}} \exp\left( r(s) + \gamma\sum_{s' \in
      \mathcal{S}} \mathcal{T}(s, a, s')V(s'; r) \right). \]
However, a reward function learned by maximising this likelihood is not
transferable to new situations
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}. One needs to model the
reward structure in a way that would allow reward predictions for previously
unseen states.

@@

% TODO: import info from the MSci paper
As training a GP with $n$ data points has a time complexity of
$\mathcal{O}(n^3)$ \cite{DBLP:books/lib/RasmussenW06}, numerous approximation
methods have been suggested, many of which select a subset of data called
\emph{inducing points} and focus most of the training effort on them
\cite{DBLP:journals/corr/abs-1807-01065}. Let $\mathbf{X_u}$ be the matrix of
features at inducing states, $\mathbf{u}$ the rewards at those states, and
$\mathbf{r}$ a vector with $r(\mathcal{S})$ as elements. Then the full joint
probability distribution can be factorised as
\begin{equation} \label{full}
  \pfull = p(\mathbf{X_u}) \times p(\Theta | \mathbf{X_u}) \times p(\mathbf{u}
  | \Theta, \mathbf{X_u}) \times p(\mathbf{r} | \Theta, \mathbf{X_u},
  \mathbf{u}) \times p(\mathcal{D} | r).
\end{equation}
Here $p(\mathbf{X_u})$ and $p(\Theta | \mathbf{X_u})$ are freely chosen priors,
\begin{align}
  p(\mathbf{u} | \Theta, \mathbf{X_u}) &= \mathcal{N}(\mathbf{u}; \mathbf{0}, \Kuu) \nonumber \\
                                       &= \frac{1}{(2\pi)^{n/2}|\Kuu|^{1/2}}\exp \left( -\frac{1}{2} \mathbf{u}^T\Kuu^{-1}\mathbf{u} \right) \nonumber \\
                                       &= \exp\left(-\frac{1}{2}\mathbf{u}^T\Kuu^{-1}\mathbf{u} - \frac{1}{2}\log|\Kuu| - \frac{m}{2}\log 2\pi\right) \label{eq:normal}
  \end{align} % TODO: introduce m here, deleting later introduction
is the GP prior \cite{DBLP:books/lib/RasmussenW06}, the GP posterior is a
multivariate Gaussian \cite{DBLP:conf/nips/LevinePK11}
\begin{equation} \label{eq:r}
  p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u}) =
  \mathcal{N}(\mathbf{r}; \Kru^T\Kuu^{-1}\mathbf{u}, \Krr - \Kru^T\Kuu^{-1}\Kru),
\end{equation}
and $p(\mathcal{D} | r)$ is as in \eqref{pDr}. The matrices such as
$\Kru$ are called \emph{covariance matrices} and are defined as
$[\Kru]_{i,j} = k_\Theta(\mathbf{x}_{\mathbf{r},i}, \mathbf{x}_{\mathbf{u},j})$,
where $\mathbf{x}_{\mathbf{r},i}$ and $\mathbf{x}_{\mathbf{u},j}$ denote feature
vectors for the $i$th state in $\mathcal{S}$ and the $j$th state in
$\mathbf{X_u}$, respectively \cite{DBLP:conf/uai/JinDAS17}.

Given this model, data $\mathcal{D}$, and inducing feature matrix
$\mathbf{X_u}$, our goal is then to find optimal values of hyperparameters
$\Theta$, inducing rewards $\mathbf{u}$, and the reward function $r$. While the
previous paper that considered this IRL model computed maximum likelihood estimates
for $\Theta$ and $\mathbf{u}$, and made an assumption that $\mathbf{r}$ in
\eqref{eq:r} has zero variance \cite{DBLP:conf/nips/LevinePK11}, we aim to
avoid this assumption and use variational inference to approximate the full
posterior distribution $\posterior$. \emph{Variational inference} (VI) is an
approximation technique for probability densities \cite{blei2017variational}.
Let $\approximation$ be our approximating family of probability distributions
for $\posterior$ with its own hyperparameter vector $\bm\nu$. Then it is up
to VI algorithms to optimise $\bm\nu$ in order to minimise the
\emph{Kullback-Leibler} (KL) divergence between the original probability
distribution and our approximation.  KL divergence (asymmetrically) measures how
different the two distributions are and in this case can be defined as follows
\cite{blei2017variational}:
\[ \DKL(\approximation || \posterior) = \Eq \left[ \log
  \frac{\approximation}{\posterior} \right]. \]
Since KL divergence is typically hard to compute, instead of minimising it, VI
typically tries to maximise the \emph{evidence lower bound} (ELBO) defined as
\cite{DBLP:books/lib/Bishop07,blei2017variational}
\[
  \begin{split}
    \mathcal{L}(\bm\nu) &= \Eq \left[ \log \frac{\pfull}{\approximation}
    \right] \\
    &= \iiint \approximation \log
    \frac{\pfull}{\approximation}\,d\Theta\,d\mathbf{u}\,d\mathbf{r}.
  \end{split}
\]

@@

Variational inference has seen a recent increase in interest among academics,
with different approaches focusing on different goals: better time complexity,
handling a wider variety of models, making approximations more accurate, and
using more complex function approximation techniques (such as neural networks)
to infer local latent variable values without having to calculate them
individually for each data point \cite{DBLP:journals/corr/abs-1711-05597}. As
our IRL model is based on a GP, we will begin by reviewing some of the VI
approaches applied specifically to GP regression. Based on a recent review of
scalable GPs \cite{DBLP:journals/corr/abs-1711-05597}, we will concentrate on
\emph{stochastic variational sparse approximations}, as they have achieved
modelling accuracy close to that of the full GP with no approximations, while
providing a time complexity of $\mathcal{O}(m^3)$, where $m$ is the number of
inducing points. Below we provide a short overview of various assumptions that
have been used in approximating \emph{sparse} GPs (i.e., GPs that use inducing
points), following on a paper by Qui\~{n}onero-Candela and Rasmussen
\cite{DBLP:journals/jmlr/CandelaR05}.

@@

\begin{table}
  \centering % TODO: recheck and explain
  \begin{tabular}{llllll} % Lower bound? Model? % TODO: comment about 'modelled' no longer needed
    \toprule
    Authors, year & Inducing points & Hyperparameters & Complexity \\
    \midrule
    Titsias, 2009 \cite{DBLP:journals/jmlr/Titsias09} & variational & variational & $\mathcal{O}(nm^2)$ \\
    Hensman et al., 2013 \cite{DBLP:conf/uai/HensmanFL13} & fixed & variational & $\mathcal{O}(m^3)$ \\
    Gal et al., 2014 \cite{DBLP:conf/nips/GalWR14} & variational & variational & $\mathcal{O}(nm^2)$ \\
    Hoang et al., 2015 \cite{DBLP:conf/icml/HoangHL15} & fixed & fixed & $\mathcal{O}(m^3)$ \\
    Cheng and Boots, 2017 \cite{DBLP:conf/nips/ChengB17} & variational & variational & $\mathcal{O}(nm_\alpha + nm_\beta^2)$ \\
    Hensman et al., 2017 \cite{DBLP:journals/jmlr/HensmanDS17} & fixed & variational & $\mathcal{O}(nm)$ \\
    Peng et al., 2017 \cite{DBLP:conf/icml/PengZZQ17} & variational & variational & $\mathcal{O}(m^3)$ \\
    \bottomrule
  \end{tabular}
  \caption{Summary of relevant VI approximations to GPs. For both inducing
    points and hyperparameters, 'fixed' means 'chosen before the algorithm
    starts', 'modelled' means 'represented by a probability distribution', and
    'variational' means 'included amongst the variational parameters'.
    Hyperparameters $m_\alpha$ and $m_\beta$ refer to the number of bases used
    to represent the GP's mean and covariance, respectively.}
  \label{table:approximations}
\end{table}

Table \ref{table:approximations} further summarises some of the recent and/or
influential GP approximation approaches that might be suitable for our GP IRL
model. In order to derive a reliable ELBO, the inducing points should be either
fixed or modelled, while hyperparameters cannot be variational (the reason will
be explained in Section \ref{sec:proposed_approach}).

@@

Since we expect $\posterior$ to be highly irregular, we would like our
approximation to be capable of representing more than just a unimodal Gaussian.
The primary way to represent complex posteriors in VI is by using
\emph{normalising flows}, i.e., a collection of invertible
functions---parametrised by additional variational parameters---that are applied
to latent variables \cite{DBLP:conf/icml/RezendeM15}. Unfortunately, this
parametrisation also means that the gradient of the joint probability
distribution w.r.t. variational parameters $\nabla_{\bm\nu}\pfull$ is no
longer zero, making an analytic expression for ELBO impossible using the usual
methods. % TODO: not full (I think) + discuss the possibility of making it
         % variational

@@

In order to properly investigate the difference between variational inference
and maximum likelihood estimation for the model, we keep other parts of the
model the same. Namely, we set the covariance function to a version of the
automatic relevance detection kernel
\cite{DBLP:conf/nips/LevinePK11,neal2012bayesian}
\[ k_\Theta(\mathbf{x}_i, \mathbf{x}_j) = \zeta\exp\left(
    -\frac{1}{2}(\mathbf{x}_i - \mathbf{x}_j)^T\bm\Lambda(\mathbf{x}_i -
    \mathbf{x}_j) - \mathbbm{1}[i \ne j]\sigma^2\Tr[\bm\Lambda] \right), \]
where $\zeta$ is the overall ``scale'' factor for how similar or distant the
states are,
\[ \bm\Lambda = \diag(\lambda_1, \dots, \lambda_d) = \diag(\bm\lambda) \]
is a diagonal matrix that determines how relevant each feature is (where $d$
denotes the number of features), $\mathbbm{1}$ is defined as
\[ \mathbbm{1}[b] = \begin{cases}
    1 & \mbox{if $b$ is true} \\
    0 & \mbox{otherwise},
  \end{cases} \]
and $\sigma^2$ is set to $10^{-2}/2$ (as the original paper noted that
the value makes little difference to the performance of the algorithm
\cite{DBLP:conf/nips/LevinePK11}). Thus we can set $\Theta = \{ \zeta,
\bm\lambda \}$. Similarly, we keep the expression for the prior of $\Theta$:
\[ p(\Theta | \mathbf{X_u}) = \exp\left( -\frac{1}{2}\Tr[\Kuu^{-2}] -
    \sum_{i=1}^d \log(\lambda_i + 1) \right). \]

Next, we can rewrite the posterior by using the chain rule and Bayes' theorem in
order to get a better sense of what we are trying to approximate:
\[
  \begin{split}
    \posterior &= p(\Theta | \mathbf{X_u}, \mathcal{D}) \times p(\mathbf{u} |
    \Theta, \mathbf{X_u}, \mathcal{D}) \times p(\mathbf{r} | \Theta,
    \mathbf{X_u}, \mathbf{u}, \mathcal{D}) \\
    &\propto p(\Theta | \mathbf{X_u}, \mathcal{D}) \times p(\mathbf{u} | \Theta,
    \mathbf{X_u}, \mathcal{D}) \times p(\mathcal{D} | r) \times p(\mathbf{r} |
    \Theta, \mathbf{X_u}, \mathbf{u}) \\
    &\propto p(\Theta | \mathbf{X_u}, \mathcal{D}) \times p(\mathcal{D} |
    \Theta, \mathbf{X_u}, \mathbf{u}) \times p(\mathbf{u} | \Theta,
    \mathbf{X_u}) \times p(\mathcal{D} | r) \times p(\mathbf{r} | \Theta,
    \mathbf{X_u}, \mathbf{u}) \\
    &\propto p(\mathcal{D} | \Theta, \mathbf{X_u}) \times p(\Theta |
    \mathbf{X_u}) \times p(\mathcal{D} | \Theta, \mathbf{X_u}, \mathbf{u})
    \times p(\mathbf{u} | \Theta, \mathbf{X_u}) \times p(\mathcal{D} | r) \times
    p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u})
  \end{split}
\]
Note that now there are only two unknown probability distributions:
$p(\mathcal{D} | \Theta, \mathbf{X_u})$ and $p(\mathcal{D} | \Theta,
\mathbf{X_u}, \mathbf{u})$, which can be computed as follows:
\begin{align*}
  p(\mathcal{D} | \Theta, \mathbf{X_u}, \mathbf{u}) = \int &p(\mathcal{D} | r) \times p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u}) \, d\mathbf{r}, \\
  p(\mathcal{D} | \Theta, \mathbf{X_u}) = \iint &p(\mathcal{D} | r) \times p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u}) \times p(\mathbf{u} | \Theta, \mathbf{X_u}) \, d\mathbf{u} \, d\mathbf{r}.
\end{align*}
This suggests the following form for the approximation:
\[ \approximation = q(\Theta) \times q(\mathbf{u} | \Theta) \times q(\mathbf{r}
  | \Theta, \mathbf{u}). \]

@@

First, as is common in the literature for applying VI to GPs
\cite{DBLP:conf/nips/ChengB17,DBLP:conf/uai/HensmanFL13,DBLP:conf/icml/HoangHL15,DBLP:journals/jmlr/Titsias09},
we simply set
\[ q(\mathbf{r} | \Theta, \mathbf{u}) = p(\mathbf{r} | \Theta, \mathbf{X_u},
  \mathbf{u}). \]
We can make a similarly justified choice for $q(\mathbf{u} | \Theta)$:
\[ q(\mathbf{u} | \Theta) = q(\mathbf{u}) = \mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S}) \]
for some variational parameters $\mathbf{m} \in \mathbb{R}^{m}$ and $\mathbf{S}
\in \mathbb{R}^{m \times m}$
\cite{DBLP:conf/nips/ChengB17,DBLP:journals/jmlr/HensmanDS17,DBLP:conf/aaai/HoangHL17}.

In order to have a reasonable way of calculating/approximating the ELBO (see
Section \ref{vi_algs}), we need to have an approximating distribution for
$\Theta$, but unfortunately all the papers in Table \ref{table:approximations}
either fix it or treat it as a variational parameter. Hence we make our first
assumption without justification from previous literature:
\[ q(\Theta) = q(\zeta) \times q(\bm\lambda). \]
As it is reasonable for $\bm\lambda$ to take any value in $\mathbb{R}^d$, we can
approximate it simply as
\[ q(\bm\lambda) = \mathcal{N}(\bm\lambda; \bm\mu, \bm\Sigma) \]
for some variational parameters $\bm\mu \in \mathbb{R}^d$ and $\bm\Sigma \in
\mathbb{R}^{d \times d}$. In contrast, as the covariance function must produce
non-negative values, we want to restrict $\zeta$ to positive real numbers. At
the same time, the distribution for $\zeta$ should have a flexible mean (i.e.,
not be tied to zero, like in the exponential distribution) and tails that
converge to zero as the value of the random variable moves away from the mean.
We might want to support some right skew, but the distribution should be close
to symmetric with at least some parameter values. This limits our choice of
distributions quite significantly, and we decide to go with the gamma
distribution as it is fairly flexible and commonly used
\cite{hogg2018introduction}. We then define the probability density function of
the gamma distribution as
\[ q(\zeta) = \Gamma(\zeta; \alpha, \beta) =
  \frac{\beta^\alpha}{\Gamma(\alpha)}\zeta^{\alpha - 1}e^{-\beta\zeta}, \]
where $\alpha > 0$ and $\beta > 0$ are two parameters of the distribution, and
$\Gamma(\cdot)$ is the gamma function. This gives us our list of variational
parameters: $\mathbf{m}, \mathbf{S}, \bm\mu, \bm\Sigma, \alpha, \beta$.

>>>>

\newcommand{\Kuu}{\mathbf{K}_{\mathbf{u},\mathbf{u}}}
\newcommand{\Krr}{\mathbf{K}_{\mathbf{r},\mathbf{r}}}
\newcommand{\Kru}{\mathbf{K}_{\mathbf{r},\mathbf{u}}}
\newcommand{\DKL}{D_{\mathrm{KL}}}
\newcommand{\pfull}{p(\mathcal{D}, \Theta, \mathbf{X_u}, \mathbf{u}, \mathbf{r})}
\newcommand{\pfullS}{p(\mathcal{D}, \Theta_s, \mathbf{X_u}, \mathbf{u}_s, \mathbf{r}_s)}
\newcommand{\posterior}{p(\Theta, \mathbf{u}, \mathbf{r} | \mathcal{D}, \mathbf{X_u})}
\newcommand{\approximation}{q_{\bm\nu}(\Theta, \mathbf{u}, \mathbf{r})}
\newcommand{\approximationS}{q_{\bm\nu}(\Theta_s, \mathbf{u}_s, \mathbf{r}_s)}
\newcommand{\Eq}{\mathbb{E}_{(\Theta, \mathbf{u}, \mathbf{r}) \sim \approximation}}

@@

The likelihood of the data can be written down as
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}
\begin{equation} \label{pDr}
  p(\mathcal{D} | r) = \prod_{i=1}^N \prod_{t=1}^T p(a_{i,t} | s_{i,t}) = \exp\left( \sum_{i=1}^N \sum_{t=1}^T Q_r(s_{i,t}, a_{i,t}) - V_r(s_{i,t}) \right),
\end{equation}
where
\begin{equation}
  Q_r(s_{i,t}, a_{i,t}) = r(s_{i,t}) + \gamma\sum_{s' \in \mathcal{S}}
  \mathcal{T}(s_{i,t}, a_{i,t}, s')V_r(s'),
\end{equation}
and $V_r(s)$ can be obtained by repeatedly applying the equation
\cite{supplementary_material}
\[ V_r(s) = \log \sum_{a \in \mathcal{A}} \exp\left( r(s) + \gamma\sum_{s' \in
      \mathcal{S}} \mathcal{T}(s, a, s')V_r(s') \right). \]
However, a reward function learned by maximising this likelihood is not
transferable to new situations
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}. One needs to model the
reward structure in a way that would allow reward predictions for previously
unseen states.

@@

% TODO: positive definite
As training a GP with $n$ data points has a time complexity of
$\mathcal{O}(n^3)$ \cite{DBLP:books/lib/RasmussenW06}, numerous approximation
methods have been suggested, many of which select a subset of data called
\emph{inducing points} and focus most of the training effort on them
\cite{DBLP:journals/corr/abs-1807-01065}. Let $\mathbf{X_u}$ be the matrix of
features at inducing states, $\mathbf{u}$ the rewards at those states, and
$\mathbf{r}$ a vector with $r(\mathcal{S})$ as elements. Then the full joint
probability distribution can be factorised as
\begin{equation} \label{full}
  \pfull = p(\mathbf{X_u}) \times p(\Theta | \mathbf{X_u}) \times p(\mathbf{u}
  | \Theta, \mathbf{X_u}) \times p(\mathbf{r} | \Theta, \mathbf{X_u},
  \mathbf{u}) \times p(\mathcal{D} | r).
\end{equation}
Here $p(\mathbf{X_u})$ and $p(\Theta | \mathbf{X_u})$ are freely chosen priors,
\begin{align}
  p(\mathbf{u} | \Theta, \mathbf{X_u}) &= \mathcal{N}(\mathbf{u}; \mathbf{0}, \Kuu) \nonumber \\
                                       &= \frac{1}{(2\pi)^{m/2}|\Kuu|^{1/2}}\exp \left( -\frac{1}{2} \mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} \right) \nonumber \\
                                       &= \exp\left(-\frac{1}{2}\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} - \frac{1}{2}\log|\Kuu| - \frac{m}{2}\log 2\pi\right) \label{eq:normal}
\end{align}
is the GP prior \cite{DBLP:books/lib/RasmussenW06}, where $m \in \mathbb{N}$ is
the number of inducing points. The GP posterior is a
multivariate Gaussian \cite{DBLP:conf/nips/LevinePK11}
\begin{equation} \label{eq:r}
  p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u}) =
  \mathcal{N}(\mathbf{r}; \Kru^\intercal\Kuu^{-1}\mathbf{u}, \Krr - \Kru^\intercal\Kuu^{-1}\Kru),
\end{equation}
and $p(\mathcal{D} | r)$ is as in \eqref{pDr}. The matrices such as
$\Kru$ are called \emph{covariance matrices} and are defined as
$[\Kru]_{i,j} = k_\Theta(\mathbf{x}_{\mathbf{r},i}, \mathbf{x}_{\mathbf{u},j})$,
where $\mathbf{x}_{\mathbf{r},i}$ and $\mathbf{x}_{\mathbf{u},j}$ denote feature
vectors for the $i$th state in $\mathcal{S}$ and the $j$th state in
$\mathbf{X_u}$, respectively \cite{DBLP:conf/uai/JinDAS17}.

Given this model, data $\mathcal{D}$, and inducing feature matrix
$\mathbf{X_u}$, our goal is then to find optimal values of hyperparameters
$\Theta$, inducing rewards $\mathbf{u}$, and the reward function $r$. While the
previous paper that considered this IRL model computed maximum likelihood estimates
for $\Theta$ and $\mathbf{u}$, and made an assumption that $\mathbf{r}$ in
\eqref{eq:r} has zero variance \cite{DBLP:conf/nips/LevinePK11}, we aim to
avoid this assumption and use variational inference to approximate the full
posterior distribution $\posterior$. \emph{Variational inference} (VI) is an
approximation technique for probability densities \cite{blei2017variational}.
Let $\approximation$ be our approximating family of probability distributions
for $\posterior$ with its own hyperparameter vector $\bm\nu$. Then it is up
to VI algorithms to optimise $\bm\nu$ in order to minimise the
\emph{Kullback-Leibler} (KL) divergence between the original probability
distribution and our approximation.  KL divergence (asymmetrically) measures how
different the two distributions are and in this case can be defined as follows
\cite{blei2017variational}:
\[ \DKL(\approximation || \posterior) = \Eq \left[ \log
  \frac{\approximation}{\posterior} \right]. \]
Since KL divergence is typically hard to compute, instead of minimising it, VI
typically tries to maximise the \emph{evidence lower bound} (ELBO) defined as
\cite{DBLP:books/lib/Bishop07,blei2017variational}
\begin{equation} \label{eq:elbo}
  \begin{split}
    \mathcal{L}(\bm\nu) &= \Eq \left[ \log \frac{\pfull}{\approximation}
    \right] \\
    &= \iiint \approximation \log
    \frac{\pfull}{\approximation}\,d\Theta\,d\mathbf{u}\,d\mathbf{r}.
  \end{split}
\end{equation}

@@

Variational inference has seen a recent increase in interest among academics,
with different approaches focusing on different goals: better time complexity,
handling a wider variety of models, making approximations more accurate, and
using more complex function approximation techniques (such as neural networks)
to infer local latent variable values without having to calculate them
individually for each data point \cite{DBLP:journals/corr/abs-1711-05597}. As
our IRL model is based on a GP, we will begin by reviewing some of the VI
approaches applied specifically to GP regression. Based on a recent review of
scalable GPs \cite{DBLP:journals/corr/abs-1711-05597}, we will concentrate on
\emph{stochastic variational sparse approximations}, as they have achieved
modelling accuracy close to that of the full GP with no approximations, while
providing a time complexity of $\mathcal{O}(m^3)$. Below we provide a short
overview of various assumptions that have been used in approximating
\emph{sparse} GPs (i.e., GPs that use inducing points), following on a paper by
Qui\~{n}onero-Candela and Rasmussen \cite{DBLP:journals/jmlr/CandelaR05}.

@@

\begin{table}
  \centering % TODO: recheck
  \begin{tabular}{llllll}
    \toprule
    Authors, year & Inducing points & Hyperparameters & Complexity \\
    \midrule
    Titsias, 2009 \cite{DBLP:journals/jmlr/Titsias09} & variational & variational & $\mathcal{O}(nm^2)$ \\
    Hensman et al., 2013 \cite{DBLP:conf/uai/HensmanFL13} & fixed & variational & $\mathcal{O}(m^3)$ \\
    Gal et al., 2014 \cite{DBLP:conf/nips/GalWR14} & variational & variational & $\mathcal{O}(nm^2)$ \\
    Hoang et al., 2015 \cite{DBLP:conf/icml/HoangHL15} & fixed & fixed & $\mathcal{O}(m^3)$ \\
    Cheng and Boots, 2017 \cite{DBLP:conf/nips/ChengB17} & variational & variational & $\mathcal{O}(nm_\alpha + nm_\beta^2)$ \\
    Hensman et al., 2017 \cite{DBLP:journals/jmlr/HensmanDS17} & fixed & variational & $\mathcal{O}(nm)$ \\
    Peng et al., 2017 \cite{DBLP:conf/icml/PengZZQ17} & variational & variational & $\mathcal{O}(m^3)$ \\
    \bottomrule
  \end{tabular}
  \caption{Summary of relevant VI approximations to GPs. For both inducing
    points and hyperparameters, 'fixed' means 'chosen before the algorithm
    starts' and 'variational' means 'included amongst the variational
    parameters'. Hyperparameters $m_\alpha$ and $m_\beta$ refer to the number of
    bases used to represent the GP's mean and covariance, respectively.}
  \label{table:approximations}
\end{table}

% TODO: expand on this
Table \ref{table:approximations} further summarises some of the recent and/or
influential GP approximation approaches that might be suitable for our GP IRL
model. In order to derive a reliable ELBO, the inducing points should be either
fixed or modelled by a probability distribution (none of the approaches do
that), while hyperparameters cannot be variational (the reason will be explained
in Section \ref{sec:proposed_approach}).

@@

Since we expect $\posterior$ to be highly irregular, we would like our
approximation to be capable of representing a wide range of possible probability
distributions. The primary way to represent complex posteriors in VI is by using
\emph{normalising flows}, i.e., a collection of invertible
functions---parametrised by additional variational parameters---that are applied
to latent variables \cite{DBLP:conf/icml/RezendeM15}. Unfortunately, this
parametrisation also means that the gradient of the joint probability
distribution w.r.t. variational parameters $\nabla_{\bm\nu}\pfull$ is no
longer zero, making an analytic expression for ELBO impossible using the usual
methods. While it may be possible to have an approximating distribution for the
flow parameters, it is uncertain how such an algorithm would behave, as it
would have to perform optimisation in a space with significantly more dimensions.

@@

In order to properly investigate the difference between variational inference
and maximum likelihood estimation for the model, we keep other parts of the
model the same. Namely, we set the covariance function to a version of the
automatic relevance detection kernel
\cite{DBLP:conf/nips/LevinePK11,neal2012bayesian}
\[ k_\Theta(\mathbf{x}_i, \mathbf{x}_j) = \zeta\exp\left(
    -\frac{1}{2}(\mathbf{x}_i - \mathbf{x}_j)^\intercal\bm\Lambda(\mathbf{x}_i -
    \mathbf{x}_j) - \mathbbm{1}[i \ne j]\sigma^2\Tr[\bm\Lambda] \right), \]
where $\zeta$ is the overall ``scale'' factor for how similar or distant the
states are,
\[ \bm\Lambda = \diag(\lambda_1, \dots, \lambda_d) = \diag(\bm\lambda) \]
is a diagonal matrix that determines how relevant each feature is (where $d$
denotes the number of features), $\mathbbm{1}$ is defined as
\[ \mathbbm{1}[b] = \begin{cases}
    1 & \mbox{if $b$ is true} \\
    0 & \mbox{otherwise},
  \end{cases} \]
and $\sigma^2$ is set to $10^{-2}/2$ (as the original paper noted that
the value makes little difference to the performance of the algorithm
\cite{DBLP:conf/nips/LevinePK11}). Thus we can set $\Theta = \{ \zeta,
\bm\lambda \}$. Similarly, we keep the expression for the prior of $\Theta$:
\begin{equation} \label{eq:theta_prior}
  p(\Theta | \mathbf{X_u}) = \exp\left( -\frac{1}{2}\Tr[\Kuu^{-2}] -
    \sum_{i=1}^d \log(\lambda_i + 1) \right).
\end{equation}

Next, we can rewrite the posterior by using the chain rule and Bayes' theorem in
order to get a better sense of what we are trying to approximate:
\[
  \begin{split}
    \posterior &= p(\Theta | \mathbf{X_u}, \mathcal{D}) \times p(\mathbf{u} |
    \Theta, \mathbf{X_u}, \mathcal{D}) \times p(\mathbf{r} | \Theta,
    \mathbf{X_u}, \mathbf{u}, \mathcal{D}) \\
    &\propto p(\Theta | \mathbf{X_u}, \mathcal{D}) \times p(\mathbf{u} | \Theta,
    \mathbf{X_u}, \mathcal{D}) \times p(\mathcal{D} | r) \times p(\mathbf{r} |
    \Theta, \mathbf{X_u}, \mathbf{u}) \\
    &\propto p(\Theta | \mathbf{X_u}, \mathcal{D}) \times p(\mathcal{D} |
    \Theta, \mathbf{X_u}, \mathbf{u}) \times p(\mathbf{u} | \Theta,
    \mathbf{X_u}) \times p(\mathcal{D} | r) \times p(\mathbf{r} | \Theta,
    \mathbf{X_u}, \mathbf{u}) \\
    &\propto p(\mathcal{D} | \Theta, \mathbf{X_u}) \times p(\Theta |
    \mathbf{X_u}) \times p(\mathcal{D} | \Theta, \mathbf{X_u}, \mathbf{u})
    \times p(\mathbf{u} | \Theta, \mathbf{X_u}) \times p(\mathcal{D} | r) \times
    p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u})
  \end{split}
\]
Note that now there are only two unknown probability distributions:
$p(\mathcal{D} | \Theta, \mathbf{X_u})$ and $p(\mathcal{D} | \Theta,
\mathbf{X_u}, \mathbf{u})$, which can be computed as follows:
\begin{align*}
  p(\mathcal{D} | \Theta, \mathbf{X_u}, \mathbf{u}) = \int &p(\mathcal{D} | r) \times p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u}) \, d\mathbf{r}, \\
  p(\mathcal{D} | \Theta, \mathbf{X_u}) = \iint &p(\mathcal{D} | r) \times p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u}) \times p(\mathbf{u} | \Theta, \mathbf{X_u}) \, d\mathbf{u} \, d\mathbf{r}.
\end{align*}
This suggests the following form for the approximation:
\begin{equation} \label{eq:approximation}
  \approximation = q(\Theta) \times q(\mathbf{u} | \Theta) \times q(\mathbf{r}
  | \Theta, \mathbf{u}).
\end{equation}

@@

First, as is common in the literature for applying VI to GPs
\cite{DBLP:conf/nips/ChengB17,DBLP:conf/uai/HensmanFL13,DBLP:conf/icml/HoangHL15,DBLP:journals/jmlr/Titsias09},
we simply set
\begin{equation}
  q(\mathbf{r} | \Theta, \mathbf{u}) = p(\mathbf{r} | \Theta, \mathbf{X_u},
  \mathbf{u}).
\end{equation}
We can make a similarly justified choice for $q(\mathbf{u} | \Theta)$:
\begin{equation}
  q(\mathbf{u} | \Theta) = q(\mathbf{u}) = \mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})
\end{equation}
for some variational parameters $\mathbf{m} \in \mathbb{R}^{m}$ and $\mathbf{S}
\in \mathbb{R}^{m \times m}$
\cite{DBLP:conf/nips/ChengB17,DBLP:journals/jmlr/HensmanDS17,DBLP:conf/aaai/HoangHL17}.

In order to have a reasonable way of calculating/approximating the ELBO (see
Section \ref{vi_algs}), we need to have an approximating distribution for
$\Theta$, but unfortunately all the papers in Table \ref{table:approximations}
either fix it or treat it as a variational parameter. Hence we make our first
assumption without justification from previous literature:
\begin{equation} \label{eq:theta}
  q(\Theta) = q(\zeta) \times q(\bm\lambda).
\end{equation}
As it is reasonable for $\bm\lambda$ to take any value in $\mathbb{R}^d$, we can
approximate it simply as
\begin{equation} \label{eq:lambda}
  q(\bm\lambda) = \mathcal{N}(\bm\lambda; \bm\mu, \bm\Sigma)
\end{equation}
for some variational parameters $\bm\mu \in \mathbb{R}^d$ and $\bm\Sigma \in
\mathbb{R}^{d \times d}$. In contrast, as the covariance function must produce
non-negative values, we want to restrict $\zeta$ to positive real numbers. At
the same time, the distribution for $\zeta$ should have a flexible mean (i.e.,
not be tied to zero, like in the exponential distribution) and tails that
converge to zero as the value of the random variable moves away from the mean.
We might want to support some right skew, but the distribution should be close
to symmetric with at least some parameter values. This limits our choice of
distributions quite significantly, and we decide to go with the gamma
distribution as it is fairly flexible and commonly used
\cite{hogg2018introduction}. We then define the probability density function of
the gamma distribution as
\begin{equation} \label{eq:zeta}
  q(\zeta) = \Gamma(\zeta; \alpha, \beta) =
  \frac{\beta^\alpha}{\Gamma(\alpha)}\zeta^{\alpha - 1}e^{-\beta\zeta},
\end{equation}
where $\alpha > 0$ and $\beta > 0$ are two parameters of the distribution, and
$\Gamma(\cdot)$ is the gamma function. This gives us our vector of variational
parameters $\bm\nu = (\mathbf{m}, \mathbf{S}, \bm\mu, \bm\Sigma, \alpha,
\beta)^\intercal$.
